

\beginedxvertical{Page One}

\beginedxtext{Preliminaries}





At the end of this sequence, and after some practice, you should be able to:

\begin{itemize}
\item Calculate the inverse of a matrix.  
\item Recognize conditions under which a matrix is invertible.   
\item Use inverses to solve matrix equations.  
\end{itemize}


For time budgeting purposes, this sequence has X videos totaling X minutes, 
plus some questions.  




\endedxtext

\endedxvertical



\beginedxvertical{What is an inverse function?}

\beginedxproblem{Cube roots}{\dpa1}

If $\ f(x) = x^3$, then here are some values of $\ f$:
\begin{align*} f(-11) &= -1331 \\ f(12) &= 1728 \\  f(-13) &= -2197. \end{align*}  

Let $g(y) = \ds{\sqrt[3]{y}}$.  What is $g(-2197)$?

\edXabox{type="numerical" expect="-13"}

\endedxproblem

\doedxvideo{Inverse Transformations and Matrices}{gv2efnioewc}


\endedxvertical



\beginedxvertical{Inverse Definitions}

\beginedxtext{Inverse Definitions}

{\keya{\bf{Definition.}}}  
If $T: V \rightarrow W$ and $S: W \rightarrow V$ are linear transformations, we say that
$S$ and $T$ are {\keyb{\bf{inverses}}} of one another if $S(T(v)) = v$ for all $v\in V$, and
$T(S(w)) = w$ for all $w\in W$.  If this is the case, we can write $S = T\inv$ and $T = S\inv$.  

If $S = T\inv$ exists we say that $T$ is {\keyb{\bf{invertible}}}.  

{\keya{\bf{Proposition.}}} A linear transformation $T$ is invertible if and only if it is both into and
onto.  

{\keya{\bf{Definition.}}}   
If $A$ is an $m\times n$ matrix and $B$ is an $n\times m$ matrix, we say that 
$A$ and $B$ are {\keyb{\bf{inverses}}} of one another if $AB = I_m$ and $BA = I_n$. If this is the case, we can write $B = A\inv$ and $B = A\inv$.  

If $B = A\inv$ exists we say that $A$ is {\keyb{\bf{invertible}}}.  



\endedxtext




\endedxvertical



\beginedxvertical{Into, Onto, and Invertibility}






\beginedxproblem{Into and Onto}{\dpa1}

We've noted that a linear transformation has an inverse if and only if it is both into and onto.  
Let's recall our results about when a linear transformation from $\R^n$ to $\R^m$ can or must
be into or onto.  

Suppose $T: \R^n \rightarrow \R^m$ is a linear transformation.


If $n<m$, what can we conclude about $T$?  

\edXabox{expect="T might be into, or might not" options="T must be into","T might be into, or might not","T cannot be into"}

\edXabox{expect="T cannot be onto" options="T must be onto","T might be onto, or might not","T cannot be onto"}


If $n>m$, what can we conclude about $T$?  

\edXabox{expect="T cannot be into" options="T must be into","T might be into, or might not","T cannot be into"}

\edXabox{expect="T might be onto, or might not" options="T must be onto","T might be onto, or might not","T cannot be onto"}



% If $n=m$, what can we conclude about $T$?  (Check all that apply.)  

% \edXabox{expect="T might be into, or might not" options="T must be into","T might be into, or might not","T cannot be into}

% \edXabox{expect="T might be onto, or might not" options="T must be onto","T might be onto, or might not","T cannot be onto}

\edXsolution{
}

\endedxproblem

\beginedxproblem{When is invertible possible?}{\dpa1}

Again, let $T: \R^n \rightarrow \R^m$ be a linear transformation.  Given the above, 
if $n<m$, is it possible for $T$ to be invertible?

\edXabox{expect="No" options="Yes","No"}

If $n>m$, is it possible for $T$ to be invertible?

\edXabox{expect="No" options="Yes","No"}


\edXsolution{
}

\endedxproblem


\endedxvertical



\beginedxvertical{Conditions for Invertibility}




\doedxvideo{Conditions for Invertibility}{pYqavDw0wt4}


\beginedxproblem{Square Matrices}{\dpa1}

True or False: If a matrix $A$ is not square, it cannot be invertible.  

\edXabox{expect="True" options="True","False"}

True or False: If a matrix $A$ is square, it must be invertible.  

\edXabox{expect="False" options="True","False"}


\edXsolution{ Watch the last video again!  Non-square matrices cannot be invertible, but not 
every square matrix is invertible.  For instance, a square matrix of all zeros is not invertible.  

}
\endedxproblem


\endedxvertical



\beginedxvertical{Invertible Matrix Theorem}


\beginedxtext{Inverse Definitions}

{\keya{\bf{Invertible Matrix Theorem.}}}  Suppose that $T: \R^n \rightarrow \R^n$ is a linear transformation
with (square) standard matrix $A$.  Then the following conditions are all equivalent (i.e., either all of
them are true, or all of them are false):

\begin{enumerate}
\item $T$ is into.
\item $\mathrm{Ker}(T) = \{ \veco \}$.  
\item The columns of $A$ are linearly independent. 
\item $Ax = \veco$ has only the trivial solution.
\item $A$ row reduces to have a pivot in every column (has no free variables).  
\item $A$ row reduces to the identity matrix.  
\item $A$ row reduces to have a pivot in every row.  
\item For every $v\in \R^n$, the equation $Ax = v$ is consistent.
\item The columns of $A$ span $\R^n$.  
\item $T$ is onto.  
\item $T$ is invertible.
\item $A$ is invertible.

\end{enumerate}

Note: There will be more equivalent conditions added to this list in the near future!  


\endedxtext



\beginedxproblem{Using the Invertible Matrix Theorem}{\dpa1}

Suppose $A$ is an $m\times n$ matrix which has at least one free variable.  Can we conclude that the columns of 
$A$ do not span $\R^m$?  

\edXabox{expect="No" options="Yes","No"}

Suppose $A$ is an $n\times n$ matrix which has at least one free variable. Can we conclude that the columns of $A$ do not span $\R^n$?  

\edXabox{expect="Yes" options="Yes","No"}

\edXsolution{ For the first question, we cannot make that conclusion, since we are not told the matrix is square.  For instance, $\left[ \begin{array}{ccc} 1 & 0 & 0 \\ 0 & 1 & 0 \end{array} \right]$ has a free variable,
but its columns span $\R^2$.  

For the second question, we know the matrix is square, so the Invertible Matrix Theorem applies.
Since $A$ has a free variable, we can conclude that none of the conditions in the Invertible Matrix Theorem
are true.  In particular, its columns do not span $\R^n$.  
}
\endedxproblem



\endedxvertical



\beginedxvertical{Calculating the Inverse}


\beginedxproblem{Is it Invertible?}{\dpa1}

Let $A = \left[ \begin{array}{cc} 1 & 2 \\ 2 & 5 \end{array} \right]$.  By row-reducing $A$, answer
the question, is $A$ invertible?

\edXabox{expect="Yes" options="Yes","No"}


\edXsolution{ $A$ row reduces to the identity matrix, so it is invertible.  
}

\endedxproblem


\doedxvideo{Calculating the Inverse}{qraxzr8nb5o}



\endedxvertical


\beginedxvertical{Inverse Practice}




\beginedxproblem{Invert a Matrix}{\dpa3}

The following $3\times 3$ matrix is invertible:

\[A = \left[ \begin{array}{ccc} 1 & 1 & 2  \\
3 & 0  & -2  \\
-1 & 1 & 0
 \end{array} \right] \]

Find $A\inv$.  


\input{matrixentry.tex}

\edXabox{type="custom" cfn="MatrixEntry" expect="[[0.2,0.2,-0.2],[0.2,0.2,0.8],[0.3,-0.2,-0.3]]"}

\edXsolution{

}

\endedxproblem


\beginedxproblem{Solve Using Inverse}{\dpa1}


Let  $A$ be the matrix from above.  Without doing any further row-reduction, find the solution
to $Ax = \left[ \begin{array}{c} 100 \\ 0 \\ 10 \end{array} \right].$

\input{vectorentry.tex}


\edXabox{type="custom" cfn="VectorEntry" expect="[[18],[28],[27]]"}

\edXsolution{ 

	
}


\endedxproblem


\endedxvertical



\beginedxvertical{One-Sided Inverses?}


\beginedxproblem{Non-square inverse? 1}{\dpa1}

Is it possible for a $2\times 3$ matrix to be invertible?  

\edXabox{expect="No" options="Yes","No"}


\edXsolution{ We have shown that non-square matrices cannot be invertible.  
}

\endedxproblem

\beginedxproblem{An Interesting Product}{\dpa3}

Given 
\[A = \left[ \begin{array}{ccc} 1 & 2 & 0  \\
0 & 1 & 1
 \end{array} \right], \  \mathrm{and} \ B = \left[ \begin{array}{cc} 1 & 0  \\
0 & 0 \\
0 & 1
 \end{array} \right], \]
calculate $AB$.  


\input{matrixentry2.tex}

\edXabox{type="custom" cfn="MatrixEntry" expect="[[1,0],[0,1]]"}

\edXsolution{

}

\endedxproblem

\beginedxproblem{Non-square inverse? 2}{\dpa1}

Look at the result of the previous problem.  
Does this mean that $A$ and $B$ are inverses of one another?

\edXabox{expect="No" options="Yes","No"}


\edXsolution{ No.  For $A$ and $B$ to be inverses, {\keyb{\bf{both}}}  $AB$ and $BA$ must be the identity.  
}

\endedxproblem


\endedxvertical


\beginedxvertical{One Side Only}

\doedxvideo{One-Sided Inverses}{X5sMleqtvAk}

\beginedxtext{One-Sided Inverses}

{\keya{\bf{Proposition.}}}  If  $A$ and $B$ are square matrices, and 
$AB = I$, then $BA = I$, and the two matrices are inverses of each other.  


This allows us to add two lines to the Invertible Matrix Theorem: 

{\keya{\bf{Invertible Matrix Theorem (updated).}}}  
Suppose that $T: \R^n \rightarrow \R^n$ is a linear transformation
with (square) standard matrix $A$.  Then the following conditions are all equivalent (i.e., either all of
them are true, or all of them are false):

\begin{enumerate}
\item $T$ is into.
\item $\mathrm{Ker}(T) = \{ \veco \}$.  
\item The columns of $A$ are linearly independent. 
\item $Ax = \veco$ has only the trivial solution.
\item $A$ row reduces to have a pivot in every column (has no free variables).  
\item $A$ row reduces to the identity matrix.  
\item $A$ row reduces to have a pivot in every row.  
\item For every $v\in \R^n$, the equation $Ax = v$ is consistent.
\item The columns of $A$ span $\R^n$.  
\item $T$ is onto.  
\item $T$ is invertible.
\item $A$ is invertible.
\item There is a matrix $B$ such that $AB = I_n$.  ($A$ has a right inverse.)
\item There is a matrix $B$ such that $BA = I_n$.  ($A$ has a left inverse.)  

\end{enumerate}



\endedxtext


\beginedxproblem{Using the IMT}{\dpa1}

Let $T: \R^3 \rightarrow \R^3$ be a linear transformation with standard matrix $A$.  Suppose the columns of $A$ are linearly
independent.  Which of the following must be true?  Click all that apply.  

\edXabox{type="oldmultichoice" expect="$T$ is onto","$A$ is invertible","The equation $Ax = \left[\begin{array}{c} 1 & 3 & 5 \end{array} \right]$ has a unique solution","$T$ is into","$A$ row reduces to the identity matrix" options="$T$ is onto","$A$ is invertible","$A$ has a free variable","The equation $Ax = \left[\begin{array}{c} 1 & 3 & 5 \end{array} \right]$ has exactly one solution","$T$ is into","$A$ row reduces to the identity matrix"}


\edXsolution{
}

\endedxproblem


\endedxvertical