

\beginedxvertical{Page One}

\beginedxtext{Preliminaries}





At the end of this sequence, and after some practice, you should be able to:

\begin{itemize}
\item Understand the relationship between elementary matrices and row operations.  
\item Find the transpose of a matrix.  
\item Recognize more conditions under which a matrix is invertible.   
\end{itemize}


For time budgeting purposes, this sequence has X videos totaling X minutes, 
plus some questions.  




\endedxtext

\endedxvertical



\beginedxvertical{Inverses and Products}


\doedxvideo{Inverses and Products}{KdbmhHwABYo}

\beginedxtext{Inverses and Products}

{\keya{\bf{Proposition.}}}  
If $A$ and $B$ are $n\times n$ matrices which are both invertible, then $AB$ is invertible, and 
$(AB)\inv = B\inv A\inv$. 


{\keya{\bf{Proposition.}}}  If $A$ and $B$ are $n\times n$ matrices, and their product $AB$ is invertible, then
both $A$ and $B$ are invertible.  

\endedxtext




\endedxvertical


\beginedxvertical{Inverse Questions}




\beginedxproblem{Invertible Product}{\dpa1}

True or false: It is possible for two non-invertible matrices to have an invertible product.  

\edXabox{expect="True" options="True","False"}

\edXsolution{ The proposition proven on the previous pages states that if the product of two {\keyb{\bf{square}}}
matrices is invertible, then each individual matrix is.  However, this problem does not indicate that the matrices
are square.  

For instance, given 
\[A = \left[ \begin{array}{ccc} 1 & 2 & 0  \\
0 & 1 & 1
 \end{array} \right], \  \mathrm{and} \ B = \left[ \begin{array}{cc} 1 & 0  \\
0 & 0 \\
0 & 1
 \end{array} \right], \]
we find that $AB$ is the $2\times 2$ identity matrix, which is invertible, yet neither $A$ nor $B$ are invertible.

 }

\endedxproblem

\beginedxproblem{Solve for A}{\dpa1}

Suppose that $A,B,C,D$ are all invertible $n\times n$ matrices such that
$DA\inv B\inv = C\inv$.  Solve for $A$.  


\edXabox{type="multichoice" expect="$A=B\inv CD$" options="$A=DCB\inv$","$A=DB\inv C$","$A=B\inv CD$","$A=BCD\inv$","$A=D\inv CB$","$A=CB\inv D$","None of the above"}

\edXsolution{ 
If we multiply on both sides of the equation by $D\inv$ from the left, we obtain
\[A\inv B\inv = D\inv C\inv.\]
Multiplying by $B$ on the right yields
\[A\inv = D\inv C\inv B.\]
Finally, we invert both sides:
\[ A =  (D\inv C\inv B)\inv = B\inv CD.\]
 }
 
\endedxproblem


\endedxvertical

\beginedxvertical{Introducing Elementary Matrices}


\beginedxproblem{Find a matrix}{\dpa3}

Find a matrix $E$ such that, for any matrix
$A =   \left[ \begin{array}{cc} a_{11} & a_{12}  \\
a_{21} & a_{22}  \\ 
a_{31} & a_{32}  
 \end{array} \right],$ we get

\[EA = \left[ \begin{array}{cc} a_{11} & a_{12}  \\
a_{21} & a_{22}  \\ 
4a_{31} & 4a_{32}  
 \end{array} \right].\]


\input{matrixentry.tex}

\edXabox{type="custom" cfn="MatrixEntry" expect="[[1,0,0],[0,1,0],[0,0,4]]"}

\edXsolution{

}

\endedxproblem


\doedxvideo{Introducing Elementary Matrices}{20gbxsXOooY}


\beginedxproblem{Find another matrix}{\dpa3}

Find a matrix $E$ such that, for any matrix
$A =   \left[ \begin{array}{cc} a_{11} & a_{12}  \\
a_{21} & a_{22}  \\ 
a_{31} & a_{32}  
 \end{array} \right],$ we get

\[EA = \left[ \begin{array}{cc} a_{21} & a_{22}  \\
a_{11} & a_{12}  \\ 
a_{31} & a_{32}  
 \end{array} \right].\]


\input{matrixentry2.tex}

\edXabox{type="custom" cfn="MatrixEntry" expect="[[0,1,0],[1,0,0],[0,0,1]]"}

\edXsolution{

}

\endedxproblem

\endedxvertical

\beginedxvertical{Elementary Matrices}



\beginedxtext{Elementary Matrices}

{\keya{\bf{Definition.}}}  An {\keyb{\bf{elementary matrix}}} is a square matrix of one of three types.  


\begin{edXshowhide}{Elementary Matrix of the First Type}
An elementary matrix of the first type is one obtained from the identity matrix by changing the diagonal entry in the 
$i$th row and column to a number 
$c$, where $c \ne 0$:  
\[ E_1 = \left[ 
\begin{array}{ccccccc}
1 &  &  &  &  &  &  \\
 & \ddots &  &  &  &  &  \\
 &   & 1& &  &  &  \\
 &   &  &  c &  &  &  \\
 &   &  &  & 1 &  &  \\
 &   &  &  &  & \ddots  &  \\
 &   &  &  &  &   & 1 
\end{array}
\right].  \]
Multiplying  a matrix $A$ by $E_1$ (where $E_1$ is on the left) has the same effect as scaling row $i$ of $A$ by the scalar $c$.  
\end{edXshowhide}

\begin{edXshowhide}{Elementary Matrix of the Second Type}
An elementary matrix of the second type is one obtained from the identity matrix by changing the non-diagonal entry in the $i$th row and $j$th column ($i\ne j$) to a number $c \ne 0$: 
\[ E_2 = \left[ 
\begin{array}{ccccccc}
1 &  &  &  &  &  &  \\
 & \ddots &  &  &  &  &  \\
 &   & 1& &  &  &  \\
 &   &  &  \ddots &  &  &  \\
 &   &  c &  & 1 &  &  \\
 &   &  &  &  & \ddots  &  \\
 &   &  &  &  &   & 1 
\end{array}
\right].  \]
Multiplying  a matrix $A$ by $E_2$ (where $E_2$ is on the left) has the same effect as adding $c$ times the $j$th row of $A$ to row $i$.  
\end{edXshowhide}

\begin{edXshowhide}{Elementary Matrix of the Third Type}
An elementary matrix of the third type is one obtained from the identity matrix by changing the $i$th and $j$th diagonal entries to 0, and changing the entries in the $i$th row and $j$th column, and the $j$th row and $i$th column, to 1: 
\[ E_3 = \left[ 
\begin{array}{ccccccccccc}
1 &  &  &  &  &  &  & & & &\\
 & \ddots &  &  &  &  &  & & & &\\
 &   & 1& &  &  &  & & & & \\
 &   &  &  0 &  &  &  & 1 & & & \\
 &   &   &  & 1 &  &  & & & &\\
 &   &  &  &  & \ddots  & & & & & \\
 &   &  &  &  &   & 1 & & & & \\
&   &  &  1 &  &   &  & 0 & & &  \\ 
&   &  &  &  &   &  &  & 1 & & \\
&   &  &  &  &   &  &  &  & \ddots & \\ 
&   &  &  &  &   &  &  &  & & 1
 \end{array}
\right].  \]
Multiplying  a matrix $A$ by $E_3$ (where $E_3$ is on the left) has the same effect as swapping the $i$th and $j$th rows
of $A$.  
\end{edXshowhide}




\endedxtext

\endedxvertical

\beginedxvertical{Elementary Matrices and Inverses}


\doedxvideo{Elementary Matrices and Inverses}{EtoZ2FyFKS4}


\endedxvertical

\beginedxvertical{Transposes}

\beginedxtext{Transposes}

The last matrix operation we will introduce in this section is the transpose.  The transpose of a matrix $A$
is essentially the matrix whose columns are the rows of $A$ (and whose rows are the columns of $A$).  Essentially,
$A$ gets flipped across its diagonal.  More formally:

{\keya{\bf{Definition.}}}  Given an $m\times n$ matrix $A$, the {\keyb{\bf{transpose}}} of $A$ is the
$n\times m$ matrix denoted $A^t$ whose entry in the $i$th row and $j$th column is the same as the
entry in the $j$th row and $i$th column of $A$.  

For instance, if \[A = \left[ \begin{array}{ccc} 1 & 2 & 0  \\
3 & 4 & -1
 \end{array} \right],\] then \[A^t  = \left[ \begin{array}{cc} 1 & 3  \\
2 & 4 \\ 0 & -1
 \end{array} \right].\]
 
\endedxtext


\beginedxproblem{Find a Transpose}{\dpa3}

Suppose $B = \left[ \begin{array}{cc} 2 & 0  \\
1 & 3 \\ -2 & -1 \\ 5 & 0
 \end{array} \right].$  
 
What is $B^t$?  

\input{matrixentry2.tex}

\edXabox{type="custom" cfn="MatrixEntry" expect="[[2, 1, -2, 5],[0,3,-1,0]]"}

\edXsolution{The first column of $B$ becomes the first row of $B^t$.  The second column of $B$
becomes the second row of $B^t$.  
}

\endedxproblem



\beginedxproblem{Double Transpose}{\dpa1}

True or false: Any matrix $A$ is equal to the transpose of its transpose.  In other words, $A = (A^t)^t$.  

\edXabox{expect="True" options="True","False"}

\edXsolution{ Switching the rows and columns of the transpose gives the original matrix.  
 }

\endedxproblem


\endedxvertical

\beginedxvertical{Transposes and Products}

\beginedxproblem{Possible Product}{\dpa1}

Suppose $A$ is a $4\times 3$ matrix, and $B$ is an $3 \times 5$ matrix, so that $AB$ is a $4\times 5$ matrix.  

Which of the following matrix products is defined?  

\edXabox{type="multichoice" expect="$B^tA^t$" options="$A^tB^t$","$B^tA^t$","Both are defined","Neither are defined"}

\edXsolution{ 
$A^t$ is a $3\times 4$ matrix, and $B^t$ is a $5 \times 3$ matrix. For the product to be defined, the first matrix must have as many columns as the second matrix has rows. This is true in one arrangement, but not the other. 
 }

\endedxproblem



\doedxvideo{Transposes and Products}{AT6y-Q3TI_w}

\endedxvertical

\beginedxvertical{Transposes and Inverses}

\beginedxproblem{Symmetric Matrices}{\dpa1}

{\keya{\bf{Definition.}}} A matrix $A$ is {\keyb{\bf{symmetric}}} if $A = A^t$.  

Of the following matrices, which ones are symmetric?  Click all that apply.  

\edXabox{type="oldmultichoice" expect="$\left[ \begin{array}{cc} \pi & 1   \\ 1 & 4 \end{array} \right]$","$I_n$" options="$\left[ \begin{array}{ccc} 1 & 0 & 0  \\ 0 & 2 & 0 \end{array} \right]$","$\left[ \begin{array}{cc} 1 & 1   \\ 2 & 2 \end{array} \right]$","$\left[ \begin{array}{cc} \pi & 1   \\ 1 & 4 \end{array} \right]$","$I_n$"}

\edXsolution{ 
 Note that only a square matrix can be symmetric.  A $2\times 3$ matrix has a transpose which is $3\times 2$,
 so it can never be equal to its transpose.  
 }

\endedxproblem

\doedxvideo{Transposes and Inverses}{bg1nsJJS-ZY}

\endedxvertical

\beginedxvertical{Summary of Results}



\beginedxtext{Transpose Results Summary}

{\keya{\bf{Proposition.}}}  If $A$ and $B$ are matrices such that the product $AB$ is defined, then 
$(AB)^t = B^t A^t$.  

{\keya{\bf{Proposition.}}}  If $A$ is invertible, then $A^t$ is invertible, and $(A^t)\inv = (A\inv)^t$.  

{\keya{\bf{Invertible Matrix Theorem (extended).}}}  For a square ($n\times n$) matrix $A$, the following statements are all
equivalent to one another, and to the conditions previously stated in the Invertible Matrix Theorem:
\begin{itemize}
\item $A$ is invertible.
\item $A^t$ is invertible.
\item The rows of $A$ (viewed as vectors in $\R^n$) span $\R^n$.  
\item The rows of $A$ are linearly independent.  
\end{itemize}

 
\endedxtext


\beginedxproblem{Solve for A, again}{\dpa1}

Suppose that $A,B,C,D$ are all invertible $n\times n$ matrices such that
$B^t (A\inv)^t = D^t C\inv$.  Solve for $A$.  


\edXabox{type="multichoice" expect="$A=BD\inv C^t$" options="$A=D\inv C^t B$","$A=B\inv D\inv C^t$","$A=BD\inv C^t$","$A=D\inv (C\inv )^t B$","$A=DC^t B\inv$","$A=C^t D\inv B\inv$","$A=BC^t D\inv $","$A=C^t D\inv B$","None of the above"}


\edXsolution{ 
One possible procedure: left-multiply on both sides of the equation by $(B^t)\inv$ to obtain 
\[(A\inv)^t = (B^t)\inv D^t C\inv .\] 
Take the transpose of both sides:
\[A\inv = ((B^t)\inv D^t C\inv)^t = (C\inv)^t D ((B^t)\inv)^t = (C^t)\inv D B\inv.\] 
Now we can take the inverse of both sides, and we get
\[A = B D\inv C^t.\]
 }
 
\endedxproblem


\beginedxproblem{More IMT}{\dpa3}

Let $T: \R^4 \rightarrow \R^4$ be a linear transformation with standard matrix $A$.  Suppose the second row of $A$ is equal
to three times the last row of $A$ minus the first row of $A$.  Which of the following must be true?  Click all that apply.  

\edXabox{type="oldmultichoice" expect="$A$ has a free variable" options="$T$ is onto","$A$ is invertible","$A$ has a free variable","The equation $Ax = \left[\begin{array}{c} 1 \\ 3 \\ 5 \end{array} \right]$ has exactly one solution","$T$ is into","$A$ row reduces to the identity matrix"}




\edXsolution{
Because one of the rows is a linear combination of the others, the rows are not linearly independent and 
the matrix is not invertible.  If we row reduce this matrix, we will not get the identity matrix, but will instead have a row of all zeros and  at least one free variable.  Therefore none of the conditions of the Invertible Matrix Theorem are satisfied:  $T$ is neither onto nor into, etc.  Since $A$ has a free variable,
no $Ax = v$ equation can have a unique solution (it might have zero solutions, or infinitely many).  
}

\endedxproblem



\endedxvertical

