

\beginedxvertical{Page One}

\beginedxtext{Preliminaries}





At the end of this sequence, and after some practice, you should be able to:

\begin{itemize}
\item Determine whether a linear transformation is into or onto;
\item Understand the connection between the kernel/image of a linear transformation, and into/onto.  
\end{itemize}


For time budgeting purposes, this sequence has X videos totaling X minutes, 
plus some questions.  




\endedxtext

\endedxvertical


\beginedxvertical{Image of a Linear Transformation}



\doedxvideo{Image of a Linear Transformation}{n3RkuIlIfjI}





\endedxvertical



\beginedxvertical{Image Definition}

\beginedxtext{Image Definition and Relation with Span}

{\keya{\bf{Definition.}}}  
Given a linear transformation $T: V\rightarrow W$, the  {\keyb{\bf image}} of $T$ is the set
\[ \mathrm{Image}(T) =  \left\{ T(v) \ : \ v \in V\right\}. \] 

{\keya{\bf{Proposition.}}}  
If $T: \R^n \rightarrow \R^m$ is a linear transformation with standard matrix $A$, then the image of $T$
is the span of the columns of $A$.  (This is sometimes also called the column space of $A$.)  

\endedxtext





\beginedxproblem{Existence vs. Uniqueness}{\dpa1}

Suppose $T: V\rightarrow W$ is a linear transformation, and let $w\in W$.  Is the question 
of whether $w$ is in the image of $T$ an existence question or a uniqueness question?

\edXabox{expect="Existence" options="Existence","Uniqueness"}

\edXsolution{This is equivalent to asking whether there exists $v\in V$ such that $T(v)=w$.
}

\endedxproblem


\beginedxproblem{In the Image?}{\dpa2}

Let $T: \R^3 \rightarrow \R^3$ be the linear transformation with standard matrix 
\[ A = \left[\begin{array}{ccc} 1 & 2 & 3 \\ 0 & 1 & 1 \\ -1 & 4 & 3  \end{array} \right]. \]

Is the vector $\left[\begin{array}{c} 1 \\ -1  \\ -7 \end{array} \right]$ in the image of $T$?

\edXabox{expect="Yes" options="Yes","No"}


Is the vector $\left[\begin{array}{c} 2 \\ -1  \\ 0 \end{array} \right]$ in the image of $T$?


\edXabox{expect="No" options="Yes","No"}

\edXsolution{The first question is equivalent to asking if there is a solution to $Ax=\left[\begin{array}{c} 1 \\ -1  \\ -7 \end{array} \right]$.  Since the augmented matrix \[ \left[\begin{array}{ccc:c} 1 & 2 & 3&1 \\ 0 & 1 & 1&-1 \\ -1 & 4 & 3 &-7 \end{array} \right]\] row reduces to \[ \left[\begin{array}{ccc:c} 1 & 0 & 1&3\\ 0 & 1& 1 &-1 \\ 0 & 0 & 0 & 0 \end{array} \right],\] we see that the system is consistent, so a solution exists.\\

The second question is equivalent to asking if there is a solution to $Ax=\left[\begin{array}{c} 2 \\ -1  \\ 0 \end{array} \right]$.  Since the augmented matrix \[ \left[\begin{array}{ccc:c} 1 & 2 & 3&2 \\ 0 & 1 & 1&-1 \\ -1 & 4 & 3 &0 \end{array} \right]\] row reduces to \[ \left[\begin{array}{ccc:c} 1 & 0 & 1&0\\ 0 & 1& 1 &0 \\ 0 & 0 & 0 & 1 \end{array} \right],\] we see that the system is inconsistent, so no solution exists.  

}
\endedxproblem


\endedxvertical


\beginedxvertical{Onto}


\doedxvideo{Onto}{i9Zg8aOTdwc}

\beginedxtext{Definition of Onto}

{\keya{\bf{Definition.}}}  
A linear transformation $T: V\rightarrow W$ is called  {\keyb{\bf onto}} if $ \mathrm{Image(T)}  = W$.  


\endedxtext



\beginedxproblem{Is it onto?}{\dpa2}

Let $T: \R^3 \rightarrow \R^3$ be the linear transformation with standard matrix 
\[ A = \left[\begin{array}{ccc} 1 & 2 & 3 \\ 0 & 1 & 1 \\ -1 & 4 & 3  \end{array} \right]. \]

(This is the same $T$ as in previous problems; you may wish to review them.)  

Is $T$ onto?  

\edXabox{expect="No" options="Yes","No"}


\edXsolution{If $T$ is onto, then every vector in $\R^3$ is in the image of $T$, but recall that we found that the vector $\left[\begin{array}{c} 2 \\ -1  \\ 0 \end{array} \right]$ is not in the image of $T$.  

}

\endedxproblem



\beginedxproblem{Row Reduction and Onto}{\dpa1}

Suppose $T: \R^n \rightarrow \R^m$ is a linear transformation with standard matrix $A$.  
$T$ is onto if and only if $A$ row reduces to have...


\edXabox{type="multichoice" expect="no rows of all zeros" options="a row of all zeros","no rows of all zeros","a free variable","no free variables"}

\edXsolution{We know that $T$ is onto if and only if the system $Ax=v$ is consistent for all $v\in\R^m$.  This is true if and only if $A$ row reduces to a matrix with no rows of all zeros.

}

\endedxproblem




\beginedxproblem{True or False}{\dpa2}

True or False: No linear transformation $T: \R^6 \rightarrow \R^8$ can be onto.  

\edXabox{expect="True" options="True","False"}

True or False: Every linear transformation $T: \R^8 \rightarrow \R^6$ must be onto.  

\edXabox{expect="False" options="True","False"}

\edXsolution{ Suppose $T: \R^6 \rightarrow \R^8$ is a linear transformation.  From the video, since $6<8$, 
$T$ cannot be onto.  
% Then $T$ is onto if and only if the columns of $A$ span $\R^8$.  But $A$ has only six columns, which cannot span $\R^8$ (to see this, 
% If we row reduce $A$, it can have at most 6 pivots.  There are 8 rows so there must a row of all zeroes.
% Therefore \\
\\
A linear transformation from $\R^8$ to $\R^6$ might be onto, but it might not be.  For instance, the map
sending everything to $\veco$ is a linear transformation which is not onto.  
% Next, suppose that $T: \R^8 \rightarrow \R^6$ is a linear transformation with standard matrix $A$.  Then $A$ is a $6\times 8$ matrix.  It may be the case that $A$ row reduces to have strictly less than 6 pivots, and thus, the columns span a proper subspace of $\R^6$.  In this case, $T$ would not be onto.
}
\endedxproblem

\endedxvertical

\beginedxvertical{Into}



\doedxvideo{Into}{4go4BarPvvY}


\endedxvertical



\beginedxvertical{Into Definition and Exercise}


\beginedxtext{Into Definition}

{\keya{\bf{Definition.}}}  
A linear transformation $T: V\rightarrow W$, is  {\keyb{\bf into}} if, for every $u, v\in V$ with $u\ne v$,
we have $T(u) \ne T(v)$.  

\endedxtext


\beginedxproblem{Is it into?}{\dpa1}

Let $V$ be a vector space over $\R$, and let $T: V\rightarrow V$ be given by $T(v) = 2v$.  

If we know that $T(u) = T(v)$, must it be true that $u = v$?  

\edXabox{expect="Yes" options="Yes","No"}

Is $T$ into?

\edXabox{expect="Yes" options="Yes","No"}

\edXsolution{ If  $T(u) = T(v)$, then $2u=2v$.  Multiplying by $\frac{1}{2}$, we see that $u=v$.\\

The fact that $T(u) = T(v)$ implies $u=v$ is equivalent to the definition of into.  
}
\endedxproblem

\endedxvertical



\beginedxvertical{Kernel}



\doedxvideo{Kernel of a Linear Transformation}{2jd2BhcEOyE}


\endedxvertical



\beginedxvertical{Kernel Definition and Questions}


\beginedxtext{Kernel Definition}

{\keya{\bf{Definition.}}}  
Given a linear transformation $T: V\rightarrow W$, the  {\keyb{\bf kernel}} of $T$ is the set
\[ \mathrm{Ker}(T) =  \left\{ v\in V \ : T(v) = \veco \right\}. \] 

{\keya{\bf{Proposition.}}}  
A linear transformation $T: V \rightarrow W$ is into if and only if $\mathrm{Ker}(T) = \{ \veco \}$.  

\endedxtext




\beginedxproblem{Existence vs. Uniqueness}{\dpa1}

Suppose $T: V\rightarrow W$ is a linear transformation.  Is the question 
of whether $T$ is into an existence question or a uniqueness question?

\edXabox{expect="Uniqueness" options="Existence","Uniqueness"}

\edXsolution{$T$ is into if, for any $v$, the equation $T(x) = v$ never has more than one solution.  This is 
a uniqueness question. 
% The question of whether $T$ is into is equivalent to asking if $\mathrm{Ker}(T) = \{ \veco \}$.  This is equivalent to asking if $\veco$ is the unique element in $\mathrm{Ker}(T)$.
}

\endedxproblem




\beginedxproblem{Is it into? 2}{\dpa2}

Let $T: \R^3 \rightarrow \R^4$ be the linear transformation with standard matrix 
\[ A = \left[\begin{array}{ccc} 1 & 2 & 5 \\ 0 & 1 & 2 \\ -1 & 1 & 1 \\ 2 & -1 & 0 \end{array} \right]. \]

How many solutions does the equation $Ax = \veco$ have?

\edXabox{expect="more than one" options="zero","one","more than one"}


Does $\mathrm{Ker}(T) = \{\veco\}$?  

\edXabox{expect="No" options="Yes","No"}


Is $T$ into?  

\edXabox{expect="No" options="Yes","No"}



\edXsolution{ Matrix $A$ row reduces to \[\left[\begin{array}{ccc} 1 & 0 & 1 \\ 0 & 1 & 2 \\ 0&0&0 \\ 0&0&0 \end{array} \right]. \]  Thus, when augmented with the zero vector, there are infinitely many solutions.  This also tells us that $\veco$ is not the only solution to $Ax=\veco$, and thus $\mathrm{Ker}(T) \neq \{\veco\}$.  
Since $T$ sends multiple vectors to $\veco$, $T$ does not satisfy the definition of into.

}
\endedxproblem



\beginedxproblem{Row Reduction and Into}{\dpa2}


Let $T: \R^n \rightarrow \R^m$ be a linear transformation with standard matrix $A$.

$T$ is into if and only if $A$ row reduces to have...

\edXabox{type="multichoice" expect="no free variables" options="a row of all zeros","no rows of all zeros","a free variable","no free variables"}

\edXsolution{ Suppose that $T$ has standard matrix $A$.  $T$ is into if and only if there is a unique solution (in particular $\veco$) to $Ax=\veco$.  This is true if and only if $A$ does not have any free variables when row reduced. 
}
\endedxproblem


\endedxvertical



\beginedxvertical{Summarizing}



\doedxvideo{Summary}{iQUlbQhB1fs}



\beginedxtext{Into Proposition}


{\keya{\bf{Proposition.}}}  
If the linear transformation $T: \R^n \rightarrow \R^m$ has standard matrix $A$, then $T$ is into if and only if $A$ has no free variables.  It is impossible for $T$ to be into if $n>m$.   

\endedxtext

\beginedxproblem{True or False 2}{\dpa2}

True or False: No linear transformation $T: \R^8 \rightarrow \R^6$ can be into.  

\edXabox{expect="True" options="True","False"}

True or False: Every linear transformation $T: \R^6 \rightarrow \R^8$ must be into.  

\edXabox{expect="False" options="True","False"}

\edXsolution{ Since $8>6$, by the proposition, a transformation $T: \R^8 \rightarrow \R^6$ cannot be into.\\

If $T: \R^6 \rightarrow \R^8$, $T$ may be into, or it may not be.  For an example of the latter, take $T$ to be the linear transformation taking every $v\in \R^6$ to the zero vector, $\veco\in \R^8$.  In this case, every vector in $\R^6$ is in the kernel of $T$, and thus $T$ is not into.
}
\endedxproblem

\endedxvertical
